Honey Bees - DESeq Interaction

### 2018-04-09 Manuscript/chapter4.tex
Examine overlap between Galbraith supplemental DEGs and our DESEq DEGs of their data
Incorporate all of Toth's comments

Although there is growing interest in how viruses and diet quality affect the health and sustainability of honeybees, as well as a recognition that such factors might operate interactively, there are only a small number of experimental studies thus far directed toward elucidating the interactive effects of these two factors in honeybees (\citealt{intNV}, \citealt{intNV2}, \citealt{intNV3}). We recently used laboratory cages and nucleus hive experiments to investigate how these two factors interact (\citealt{adamInt}). As part of our experiment, we specifically studied the interactive effects of IAPV infection and monofloral diet quality on titers and honeybee mortality rates.

@@

Consequently, in our prior study, for our nutrition factor, we examined two monofloral pollen diets, Cistus (Rockrose) and Castanea (Chestnut). Cistus pollen is generally considered less nutritious than Castanea pollen due to its lower levels of protein, amino acids, antioxidants, calcium, and iron (\citealt{DiPasquale}, \citealt{adamInt}). For our virus factor, one level contained bees that were infected with IAPV and another level contained bees that were not infected with IAPV. This experimental design resulted in four treatment groups that allowed us to assess main effects and interactive effects between diet quality and IAPV infection in honeybees.

Mortality rates of honeybees 72 hour post-inoculation differed (or did not?) among the treatment groups (mixed model ANOVA across all treatment groups, df=@@@, @@@; F=@@@; p<@@@). The effect of virus treatment (mixed model ANOVA, df=@@@, @@@; F=@@@; p<@@@), diet treatment (mixed model ANOVA, df=@@@,@@@; F=@@@; p<@@@), and interaction between the two factors (mixed model ANOVA, df=@@@, @@@; F=@@@, p=@@@) did (or did not?) differ. The virus treatment was significant: For a given diet, honeybees exposed to the virus showed significantly higher mortality rate than honeybees not exposed to the virus (Tukey HSD, p<0.05). Without virus exposure, there was only an intermediate reduction in mortality rate for bees fed Castanea pollen? (Tukey HSD, p>0.05). However, with virus exposure, there was a significant reduction in mortality rate for beeds fed Castanea pollen (Tukey HSD, p<0.05). Overall, we discovered that the higher-quality Castanea diet had the ability to significantly reduce mortality in the presence of IAPV infection compared to the lower-quality Cistus diet (Figure \ref{fig:mortalityAll}).

Comment on (Figure \ref{fig:IAPVAll}) @@@

\begin{figure}[H]
\centering
  \begin{framed}
  \includegraphics[width=0.8\textwidth]{Images/mortalityAll}
  \end{framed}
  \caption{Mortality rates between the four treatments. Each treatment consisted of 15 cages each with 35 honeybees.}
  \label{fig:mortalityAll}
\end{figure}

Comments on (Figure \ref{fig:IAPVAll}) and (Figure \ref{fig:IAPVRNASeq}). @@@

\begin{figure}[H]
\centering
  \begin{framed}
  \includegraphics[width=0.8\textwidth]{Images/IAPVAll}
  \end{framed}
  \caption{IAPV titers between the four treatments. Each treatment contains 15 samples (cages).}
  \label{fig:IAPVAll}
\end{figure}

Following up on these phenotypic findings from our previous study, we now aim to understand the corresponding underlying mechanisms. Transcriptomics is one means to achieve this goal. As it stands, there are only a small number of published experiments examining gene expression patterns related to diet effects (\citealt{alaux2}) and IAPV infection effects (\citealt{galbraith}) in honeybees. As far as we know, there are little to no studies investigating honeybee gene expression patterns specifically related to monofloral diets, and litte to no studies investigating gene expression patterns related to the interaction effects of diet in any broad sense and viral innoculation in any broad sense in honeybees.

In this study, we examine how monofloral diets and viral innoculation influence gene expression patterns in honeybees by focusing on four treatment groups (Cistus diet without IAPV exposure, Castanea diet without IAPV exposure, Cistus diet with IAPV exposure, and Castanea diet with IAPV exposure). We conduct RNA-sequencing analysis on a randomly selected subset of the honeybees we used in our previous study (as is further described in our methods section). We then examine pairwise combinations of treatment groups, the main effect of monofloral diet, the main effect of IAPV exposure, and the interactive effect of the two factors on gene expression patterns.

@@

The procedures we used to prepare virus inoculum, infect and feed caged honeybees, and quantify IAPV can be reviewed in our previous work (\citealt{adamInt}). The statistical analysis we used to study the main and interaction effects of the two factors on mortality and IAPV titers is also described in our earlier report (\citealt{adamInt}).

Fifteen cages per treatment were originally sampled. Six live honeybees from each cage were randomly selected 36 hours post inoculation and placed into tubes. Tubes were kept on dry ice and then transferred into a -80C freezer until processing. Eight cages were randomly selected from the original 15 cages, and 2 honeybees per cage were randomly selected from the original six live honeybees per cage. RNA from each pool of two honeybees were extracted using Qiagen RNeasy MiniKit followed by Qiagen DNase treatment. Samples were suspended in water to 200-400 ng/$\mu$l. All samples were then tested on a Bioanalyzer at the DNA core facility to ensure quality (RIN>8).

@@@ Which body tissue was used for RNA extraction?

Samples were sequenced starting on January 14, 2016 at the Iowa State University DNA Facility (Platform: Illumina HiSeq Sequencing; Category: Single End 100 cycle sequencing). Reads were aligned to the BeeBase Version 3.2 genome (\citealt{hbGenome}) from the Hymenoptera Genome Database (\citealt{hymenopteraDB}) using the programs GMAP and GSNAP (\citealt{gsnap}). We tested all six pairwise combinations of treatments for DEGs (pairwise DEGs). We also tested the diet main effect (diet DEGs), virus main effect (virus DEGs), and interaction term for DEGs (interaction DEGs). We then also tested for virus main effect DEGs (virus DEGs) in public data derived from a previous study exploring the gene expression of IAPV virus infection in honeybees (\citealt{galbraith}). We tested each DEG analysis using recommended parameters with DESeq2 (\citealt{deseq2}), edgeR (\citealt{edger}), and LimmaVoom (\citealt{limma}). In all cases, we used a false discovery rate (FDR) threshold of 0.05 (\citealt{benjamini}). Fisher's exact test was used to determine significant overlaps between DEG sets (whether from the same dataset but across different analysis pipelines or from different datasets across the same analysis pipelines).

@@@ What percent of reads mapped?

@@

When statistically analyzing the subset of this data that was used for RNA-sequencing analysis, we found that the mortality rates across diet quality and virus exposure at least numerically retained the same trends (Figure \ref{fig:mortalityRNASeq}).

Comments on (Figure \ref{fig:mortalityRNASeq}) and (Figure \ref{fig:IAPVRNASeq}). @@@

\begin{figure}[H]
\centering
  \begin{framed}
  \includegraphics[width=0.8\textwidth]{Images/mortalityRNASeq}
  \end{framed}
  \caption{Mortality rates between the four treatments. Each treatment here only contains the 6 samples (cages) that were used for RNA-sequencing out of its original 15 samples (cages). Two honeybees were randomly selected from each sample (cage) to create a pooled RNA-seqeuncing sample.}
  \label{fig:mortalityRNASeq}
\end{figure}

\begin{figure}[H]
\centering
  \begin{framed}
  \includegraphics[width=0.8\textwidth]{Images/IAPVRNASeq}
  \end{framed}
  \caption{IAPV titers between the four treatments. Each treatment here only contains the 6 samples (cages) that were used for RNA-sequencing out of its original 15 samples (cages). Two honeybees were randomly selected from each sample (cage) to create a pooled RNA-seqeuncing sample.}
  \label{fig:IAPVRNASeq}
\end{figure}

>>>>

Although there is growing interest in how viruses and diet quality affect the health and sustainability of honeybees, as well as a recognition that such factors might operate interactively, there are only a small number of experimental studies thus far directed toward elucidating the interactive effects of these two factors in honeybees (\citealt{intNV}, \citealt{intNV2}, \citealt{intNV3}). We recently used laboratory cages and nucleus hive experiments to investigate the health effects of these two factors, and our results show a significant interaction between diet quality and virus infection. Specifically, high quality pollen is able to mitigate virus-induced mortality to the level of diverse, polyfloral pollen (\citealt{adamInt}).

Following up on these phenotypic findings from our previous study, we now aim to understand the corresponding underlying mechanisms. Transcriptomics is one means to achieve this goal. As it stands, there are only a small number of published experiments examining gene expression patterns related to diet effects (\citealt{alaux2}) and IAPV infection effects (\citealt{galbraith}) in honeybees. As far as we know, there are few to no studies investigating honeybee gene expression patterns specifically related to monofloral diets, and few to no studies investigating gene expression patterns related to the interaction effects of diet in any broad sense and viral innoculation in any broad sense in honeybees.

In this study, we examine how monofloral diets and viral innoculation influence gene expression patterns in honeybees by focusing on four treatment groups (low quality diet without IAPV exposure, high quality diet without IAPV exposure, low quality diet with IAPV exposure, and high quality diet with IAPV exposure). We conduct RNA-sequencing analysis on a randomly selected subset of the honeybees we used in our previous study (as is further described in our methods section). We then examine pairwise combinations of treatment groups, the main effect of monofloral diet, the main effect of IAPV exposure, and the interactive effect of the two factors on gene expression patterns.

We also compare the main effect of IAPV exposure in our dataset to that obtained in a previous study conducted by Galbraith and colleagues (\citealt{galbraith}). As RNA-sequencing data can be highly noisy, this comparison allowed us to characterize how repeatable and robust our RNA-seq results were in comparison to previous studies. Importantly, we use an in-depth data visualization approach to explore and validate our data, and suggest such an approach can be useful for cross-study comparisons of RNA-sequencing data in the future.

Details of the procedures we used to prepare virus inoculum, infect and feed caged honeybees, and quantify IAPV can be reviewed in our previous work (\citealt{adamInt}). The statistical analysis we used to study the main and interaction effects of the two factors on mortality and IAPV titers is also described in our earlier report (\citealt{adamInt}).

\subsection{Design of diet experiment}

@@

Consequently, in our prior study, for our nutrition factor, we examined two monofloral pollen diets, Cistus (Rockrose) and Castanea (Chestnut). Cistus pollen is generally considered less nutritious than Castanea pollen due to its lower levels of protein, amino acids, antioxidants, calcium, and iron (\citealt{DiPasquale}, \citealt{adamInt}). For our virus factor, one level contained bees that were infected with IAPV and another level contained bees that were not infected with IAPV. This experimental design resulted in four treatment groups (Cistus pollen without IAPV exposure, Castanea pollen without IAPV exposure, Cistus pollen with IAPV exposure, and Castanea pollen with IAPV exposure) that allowed us to assess main effects and interactive effects between diet quality and IAPV infection in honeybees.

Fifteen cages per treatment were originally sampled. Six live honeybees from each cage were randomly selected 36 hours post inoculation and placed into tubes. Tubes were kept on dry ice and then transferred into a -80C freezer until processing. Eight cages were randomly selected from the original 15 cages, and 2 honeybees per cage were randomly selected from the original six live honeybees per cage. Whole body RNA from each pool of two honeybees were extracted using Qiagen RNeasy MiniKit followed by Qiagen DNase treatment. Samples were suspended in water to 200-400 ng/$\mu$l. All samples were then tested on a Bioanalyzer at the DNA core facility to ensure quality (RIN>8).

Samples were sequenced starting on January 14, 2016 at the Iowa State University DNA Facility (Platform: Illumina HiSeq Sequencing; Category: Single End 100 cycle sequencing). A standard Illumina mRNA library was prepared by the DNA facility. Reads were aligned to the BeeBase Version 3.2 genome (\citealt{hbGenome}) from the Hymenoptera Genome Database (\citealt{hymenopteraDB}) using the programs GMAP and GSNAP (\citealt{gsnap}). We tested all six pairwise combinations of treatments for DEGs (pairwise DEGs). We also tested the diet main effect (diet DEGs), virus main effect (virus DEGs), and interaction term for DEGs (interaction DEGs). We then also tested for virus main effect DEGs (virus DEGs) in public data derived from a previous study exploring the gene expression of IAPV virus infection in honeybees (\citealt{galbraith}). We tested each DEG analysis using recommended parameters with DESeq2 (\citealt{deseq2}), edgeR (\citealt{edger}), and LimmaVoom (\citealt{limma}). In all cases, we used a false discovery rate (FDR) threshold of 0.05 (\citealt{benjamini}). Fisher's exact test was used to determine significant overlaps between DEG sets (whether from the same dataset but across different analysis pipelines or from different datasets across the same analysis pipelines).

@@@ What percent of reads mapped?
@@@ Total number of raw reads
@@@ How many lanes
@@@ How many samples per lane

\subsection{Comparison to previous studies on transcriptomic response to viral infection}

@@

@@

\subsection{Phenotypic results}

We reanalyzed our previously published dataset with a subset more relevant to our RNA-sequencing approaches in the current study that have a more focused question regarding diet quality. We briefly show it again here to inform the RNA-seq comparison because we reduced the number of treatments (from eight to four) from the original published data (\citealt{adamInt}). When statistically analyzing the subset of this data that was used for RNA-sequencing analysis, we found that the mortality rates across diet quality and virus exposure at least numerically retained the same trends (Figure \ref{fig:4panels}).

Mortality rates of honeybees 72 hour post-inoculation differed (or did not?) among the treatment groups (mixed model ANOVA across all treatment groups, df=@@@, @@@; F=@@@; p<@@@). The effect of virus treatment (mixed model ANOVA, df=@@@, @@@; F=@@@; p<@@@), diet treatment (mixed model ANOVA, df=@@@,@@@; F=@@@; p<@@@), and interaction between the two factors (mixed model ANOVA, df=@@@, @@@; F=@@@, p=@@@) did (or did not?) differ. The virus treatment was significant: For a given diet, honeybees exposed to the virus showed significantly higher mortality rate than honeybees not exposed to the virus (Tukey HSD, p<0.05). Without virus exposure, there was only an intermediate reduction in mortality rate for bees fed Castanea pollen? (Tukey HSD, p>0.05). However, with virus exposure, there was a significant reduction in mortality rate for beeds fed Castanea pollen (Tukey HSD, p<0.05). Overall, we discovered that the higher-quality Castanea diet had the ability to significantly reduce mortality in the presence of IAPV infection compared to the lower-quality Cistus diet (Figure \ref{fig:4panels}).

Comment on (Figure \ref{fig:4panels}) @@@

\begin{figure}[H]
\centering
  \begin{framed}
  \includegraphics[width=0.8\textwidth]{Images/4panels}
  \end{framed}
  \caption{Mortality rates for all cages, mortality rates for subset of honeybees used for RNA-seq, IAPV titers for all cages, IAPV titers for subset of honeybees used for RNA-seq subset.}
  \label{fig:4panels}
\end{figure}

### 2018-04-09 NC_NR_VC_VR/DESeq2/Interaction/makeDataMetrics-Test.Rmd
Incorporate all of Toth's comments

---
title: "Honey Bees - DESeq"
author: Lindsay Rutter
output:
  packagedocs::package_docs:
    toc: true
    toc_collapse: true
vignette: >
  %\VignetteEngine{packagedocs::redirect}
---

@@

```{r}
library(DESeq2)
library(RColorBrewer)
library(data.table)
library(ggplot2)
library(gplots)

@@

dds = DESeqDataSetFromMatrix(countData = data, colData = coldata, design = ~ treatment)
dds <- DESeq(dds)
```

We can examine pairwise combinations of all samples

```{r}
uTreat <- unique(unlist(lapply(colnames(data), function (x) unlist(strsplit(x, "[.]"))[1])))
allPairs = data.frame(Treatment1 = factor(), Treatment2 = factor(), NumberDEG = numeric())
dataMetrics <- list()

for (i in 1:(length(uTreat)-1)){
  for (j in (i+1):length(uTreat)){
    res <- results(dds, contrast=c("treatment",uTreat[i],uTreat[j]))
    hist(res$pvalue, breaks=20, col="grey", main=paste("pvalue for", uTreat[i], "and", uTreat[j]))
    degLength <- length(which((res@listData)$padj <0.05))

    metrics = as.data.frame(res@listData)
    metrics = cbind(ID = res@rownames, metrics)
    metrics$ID = as.character(metrics$ID)
    dataMetrics[[paste0(uTreat[i], "_", uTreat[j])]] <- metrics

    t1Length <- length(which(metrics[which(metrics$padj <0.05),]$log2FoldChange>0))
    t2Length <- length(which(metrics[which(metrics$padj <0.05),]$log2FoldChange<0))

    allPairs = rbind(allPairs, data.frame(Treatment1 = uTreat[i], Treatment2 = uTreat[j], NumberDEG = degLength, Treatment1Larger = t1Length, Treatment2Larger = t2Length))
  }
}
allPairs <- allPairs[order(allPairs$NumberDEG),]
allPairs
saveRDS(dataMetrics, file="dataMetrics.Rds")
```

```{r}
plotDispEsts(dds, ylim = c(1e-6, 1e1))
```

Taking rlog transformation for EDA.

```{r}
rld <- rlog(dds)
rldDF <- as.data.frame(assay(rld))
saveRDS(rldDF, file="dataRLD.Rds")
```

```{r}
sampleDists <- dist(t(assay(rld)))
sampleDistMatrix <- as.matrix(sampleDists)
rownames(sampleDistMatrix) <- paste(rld$treatment, rld$patient, sep="-")
colnames(sampleDistMatrix) <- NULL
colours = colorRampPalette(rev(brewer.pal(9, "Blues")))(255)
heatmap.2(sampleDistMatrix, trace="none", col=colours)
```

```{r}
print(plotPCA(rld, intgroup="treatment") + geom_label(aes(label = name), label.size=0.001))
```

>>>>

---
title: "Honey Bees - DESeq Interaction"
author: Lindsay Rutter
output:
  packagedocs::package_docs:
    toc: true
    toc_collapse: true
vignette: >
  %\VignetteEngine{packagedocs::redirect}
---

@@

```{r}
library(DESeq2)
library(gplots)
library(RColorBrewer)
library(data.table)
library(ggplot2)

@@

dds = DESeqDataSetFromMatrix(countData = data, colData = coldata, design = ~ virus + diet + virus*diet)

dds <- DESeq(dds)

# Intercept, virus_V_vs_N, diet_R_vs_C, virusV.dietR
resultsNames(dds)

res <- results(dds, name="Intercept")
length(which((res@listData)$padj <0.05))

res <- results(dds, name="virus_V_vs_N")
length(which((res@listData)$padj <0.05))

res <- results(dds, name="diet_R_vs_C")
length(which((res@listData)$padj <0.05))

res <- results(dds, name="virusV.dietR")
length(which((res@listData)$padj <0.05))

```
